---
title: "Of mice and men"
format: html
editor: source
toc: true
number-sections: true
echo: true
code-fold: false
warning: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo       = TRUE,
                      message    = FALSE,
                      cache      = TRUE,
                      autodep    = TRUE,
#                      dev.args   = list(pointsize = 11),
                      fig.height = 5,
                      fig.width  = 1.4 * 5,
                      fig.retina = 2,
                      fig.align  = "center")
```

All the data and source code files are [here](https://github.com/TB-DURATION/mice).

You can ask for additional analysis [here](https://github.com/TB-DURATION/mice/issues).

## Introduction

Here we are working on the data from
Zhang *et al.*
"Short-course chemotherapy with TMC207 and rifapentine in a murine model of
latent tuberculosis infection".
*American Journal of Respiratory and Critical Care Medicine* 184.6 (2011): 732-737.
DOI: [10.1164/rccm.201103-0397OC](https://doi.org/10.1164/rccm.201103-0397oc).
Specifically the data from Table 2 shown in @fig-table-2 below.

![Table 2 from [10.1164/rccm.201103-0397OC](https://doi.org/10.1164/rccm.201103-0397oc).](table 2.png){#fig-table-2}

The data show the proportion of positive mice as a function of time (about 15
mice per time point and treatment) but what we are interested in is the CFU
clearance dynamics at the individual level. Since the former is a consequence of
the latter, we develop a simple model of CFU clearance at the individual level
from which we compute the consequences at (small-size) population level. By
confronting the model-predicted consequences at population level with actual
experimental data, we are able to estimate the two parameters of the model. Once
the model is calibrated, we can use it to explore what would be the effect of
decreasing the initial value of CFU on the time at which the individual clear
the infection.

The model itself is a simple exponential decay model with inter-individual
variability on the rate of decay that is modelled by a Gamma distribution.


## Packages

Loading the packages needed for the analysis:

```{r}
library(readxl)
library(dplyr)
library(purrr)
library(tidyr)
library(bbmle)
```


## Utilitary functions

Width of the lines:

```{r}
line_width <- 2
```

A function that adds proportions estimates and confidence intervals to a data
frame from a column of successes and a column or trials:

```{r}
add_prop_est <- function(df, x, n, p = "est", l = "lwr", u = "upr", ...) {
  df |> 
    mutate(test   = map2({{ x }}, {{ n }}, prop.test, ...),
           "{p}" := map_dbl(test, ~ .x[["estimate"]]),
           conf   = map(test, ~ setNames(.x[["conf.int"]], c(l, u)))) |> 
    unnest_wider(conf) |> 
    select(- test)
}
```

A reparameterization of `dgamma()`:

```{r}
dgamma2 <- function(x, mu, sigma) {
  dgamma(x, mu^2 / sigma^2, scale = sigma^2 / mu)
}
```

A reparameterization of `qgamma()`:

```{r}
qgamma2 <- function(p, mu, sigma) {
  qgamma(p, mu^2 / sigma^2, scale = sigma^2 / mu)
}
```

A function that draw a plot frame:

```{r}
plot_frame <- function(t_max = 6 * 30, agg = 7, log = FALSE, ...) {
  if (log) {
    plot(NA, xlim = c(0, t_max / agg), ylab = "CFU", axes = FALSE,
         ylim = c(1, 1e4), log = "y", ...)
    abline2(v = 0:30)
    abline2(h = unlist(map(10^(0:3), ~ .x * 1:10)))
  } else {
    plot(NA, xlim = c(0, t_max / agg), ylab = "CFU", axes = FALSE,
         ylim = c(0, 1e4), ...)
    abline2(v = 0:30)
    abline2(h = 0:10 * 1e3)
  }
  axis(1); axis(2)
}
```

A function that adds legend:

```{r}
add_legend <- function(where = "topright", ...) {
  legend(where, legend = c("RIF", "RIF+INH", "RPT+INH"),
         col = col_treatments[-1], lwd = line_width, ...)
} 
```

Tuning `legend()`:

```{r}
legend2 <- function(...) legend(..., bty = "n")
```

Tuning `abline()`:

```{r}
abline2 <- function(...) abline(..., col = "antiquewhite")
```

Tuning `point()`:

```{r}
points2 <- function(...) points(..., lwd = line_width)
```

Tuning `arrows()`:

```{r}
arrows2 <- function(...) {
  arrows(..., length = 0.1, angle = 90, code = 3, lwd = line_width)
}
```

Tuning `lines()`:

```{r}
lines2 <- function(...) lines(..., lwd = line_width)
```

Tuning `polygon()`:

```{r}
polygon2 <- function(x, y1, y2, col, ...) {
  polygon(c(x, rev(x)), c(y1, rev(y2)), border = NA, col = adjustcolor(col, .2), ...)
}
```

Tuning `seq()`:

```{r}
seq2 <- function(...) seq(..., le = 100)
seq3 <- function(...) seq(..., le = 512)
```

Tuning `title()`:

```{r}
title2 <- function(...) title(..., line = .5)
title3 <- function(...) title(..., line = 0)
```


## The data

Reading the data:

```{r}
table2 <- read_excel("table 2 Zhang et al 2011.xlsx")
```

The data look like this:

```{r}
table2
```

Preparing the data for vizualization by:

* computing the proportions of positive mice with 95% confidence intervals
* jittering the duration variable slightly in order to avoid visual overlap

```{r}
table2a <- table2 |>
  add_prop_est(positive, total) |> 
  mutate_at("duration", jitter, factor = .2)
```

It looks like this:

```{r}
table2a
```

The treatments:

```{r}
treatments <- unique(table2$group)
```

The colors of the treatments:

```{r}
col_treatments <- setNames(seq_along(treatments), treatments)
```

A function that plots proportion estimates and confidence intervals for a given
treatment `x`:

```{r plot_prop_est}
plot_prop_est <- function(x, col, polyg = TRUE, connect = TRUE) {
  with(x, {
    if (polyg) polygon2(duration, lwr, upr, col)
    points2(duration, est, col = col, type = ifelse(connect, "o", "p"))
    arrows2(duration, lwr, duration, upr, col = col)
    if (connect) {
      lines2(duration, lwr, lty = 2, col = col)
      lines2(duration, upr, lty = 2, col = col)
    }
  })
}
```

A function that plots the proportions estimates from the data:

```{r plot_data_prop_est, dependson = "plot_prop_est"}
plot_data_prop_est <- function(polyg = TRUE, connect = TRUE) {
  plot(NA, xlim = c(0, 6), ylim = 0:1,
     xlab = "duration of treatment (months)",
     ylab = "proportion of positive mice")

  table2a |> 
    group_by(group) |> 
    group_split() |> 
    walk2(col_treatments, plot_prop_est, polyg, connect)
}
```

The plot showing points estimates with confidence intervals from the data:

```{r dependson = "plot_data_prop_est"}
#| label: fig-data-table2
#| fig-cap: "experimental proportions of positive ($n$ = 15) for the 4 treatments of
#| Table 2 of [10.1164/rccm.201103-0397OC](https://doi.org/10.1164/rccm.201103-0397oc)."

plot_data_prop_est()
legend2("bottomright", legend = treatments, col = col_treatments, lwd = line_width)
```

## A model of CFU decay

### Exponential decay

Let's assume that the CFU in an individual decays at a constant rate $\lambda$.
Then, the CFU as a function of time $t$ reads:

$$
\mbox{CFU}(t) = \mbox{CFU}_0 e^{-\lambda t}
$$

A function that gives CFU as a function of initial CFU$_0$, decay rate and time:

```{r}
cfu <- function(cfu0, lambda, t) {
  cfu0 * exp(- lambda * t)
}
```

Let's try it:

```{r}
cfu(1e4, .115, 30)
```

### Individual variability

The next step is to consider that the rate of exponential decay varies slightly
from individual to individual. Let's consider that this inter-individual
variability follows a Gamma distribution:

$$
\lambda \sim \mbox{Gamma}(\mu,\sigma)
$$
where $\mu$ and $\sigma$ are the mean and standard deviation of the Gamma
distribution. Next, we want to work out how these $\mu$ and $\sigma$ parameters
translate into a proportion of positive individuals in a population of an
infinite number of individuals at time $t$ (positivity being defined as CFU
$\ge 1$). This is done by looking at the cumulative probability of the Gamma
distribution of the decay rate that leads to the quantile of the
inter-individual CFU distribution that is the closest to 1. To do so, we need a
function that expresses the distance of the quantile of the inter-individual CFU
distribution to 1:

```{r}
dist_to_1 <- function(p, mu, sigma, N0, t) {
  abs(1 - cfu(N0, qgamma2(p, mu, sigma), t))
}
```

We then use this function in the following function that converts values for
$\mu$ and $\sigma$ of the Gamma distribution of the inter-individual
variability of exponential decay rate into the probability parameter of a
binomial distribution that corresponds to the expected proportion of positive
individuals in the population:

```{r}
probability <- function(mu, sigma, N0 = 1e4, t, epsilon = 1e-16) {
  upr <- cfu(N0, qgamma2(epsilon, mu, sigma), t)
  if (upr < 1) {
    return(0)
  } else {
    lwr <- cfu(N0, qgamma2(1 - epsilon, mu, sigma), t)
    if (lwr > 1) {
      return(1)
    }
    else {
      optimize(dist_to_1, 0:1, mu = mu, sigma = sigma, N0 = N0, t = t)$minimum
    }
  }
}
```

Let's try it with this function:

```{r fig.height = .7 * 2 * 5, fig.width  = .7 * 1.4 * 5}
plot_cfu <- function(mu_val, sigma_val, t_val, col_val, leg) {
  plot(NA, xlim = c(0, .5), ylim = 0:1,
       xlab = expression(paste("within-host mean decay rate ", mu, " (/day)")),
       ylab = "in infinite population",
       main = paste("t =", t_val))
  
  mtext("proportion positive", 2, 2.5)

  walk2(sigma_val, col_val,
        ~ lines2(mu_val,
                 map_dbl(mu_val, probability, sigma = .x, t = t_val),
                 col = .y))
  
  if (leg) {
      legend2("bottomleft", col = col_val, lwd = line_width,
              legend = sapply(sigma_val,
                              function(x)
                                as.expression(substitute(sigma == A,
                                                         list(A = as.name(x))))),
              title = expression(paste("ind. variability ", sigma, " (/day):")))
  }
}
```

Let's look at 2 time points:

```{r fig.height = 4, fig.width  = 1.6 * 5}
#| label: fig-model-predictions
#| fig-cap: "model-predicted proportion of positive in an population of infinite size as
#| a function of (a) mean decay rate ($x$-axis), (b) inter-individual variability
#| (3 lines), and (c) time (2 panels)."

opar <- par(mfrow = 1:2, cex = 1, plt = c(.2, .95, .2, .8))

walk2(c(25, 45),
      c(TRUE, FALSE),
      ~ plot_cfu(mu_val    = seq3(.001, .5), # (/day)
                 sigma_val = c(.01, .05, .11), # (/day)
                 t_val     = .x, # (day)
                 col_val   = 2:4,
                 leg       = .y))

par(opar)
```


## ML estimation

### Likelihood function

The first step consists in expressing a function that computes minus log
likelihood:

```{r}
threshold <- - log(1e-16)

mLL <- function(mu, sigma, N0, data, epsilon = 1e-16) {
  min(threshold,
      - sum(dbinom(data$positive,
                   data$total,
                   map_dbl(data$duration,
                           ~ probability(mu, sigma, N0, .x * 30, epsilon)),
                   TRUE)))
}
```

Let's try it:

```{r}
mLL(mu = .3, sigma = .01, N0 = 1e4, data = filter(table2, group == "RIF"))
```

### Likelihood surfaces

Since we have only 2 parameters, we can easily have a look at the likelihood
surface. A function that calculates the minus log-likelihood values:

```{r calc_mLL}
calc_mLL <- function(drug, mu_val, sigma_val, N0 = 1e4, epsilon = 1e-16) {
  mLL_val <- expand.grid(mu_val, sigma_val) |>
    with(map2_dbl(Var1, Var2, mLL, N0 = N0, data = filter(table2, group == drug),
                  epsilon = epsilon)) |> 
    matrix(length(mu_val))
  list(mu_val = mu_val, sigma_val = sigma_val, mLL_val = mLL_val)
}
```

Calculating the minus log-likelihood values:

```{r dependson = "calc_mLL"}
mLL_INH <- calc_mLL("INH", seq2(.001, .05), seq2(.0001, .012))
mLL_RIF <- calc_mLL("RIF", seq2(.001, .2), seq2(.0001, .1))
mLL_RIFINH <- calc_mLL("RIF+INH", seq2(.05, .15), seq2(.0001, .15))
mLL_RPTINH <- calc_mLL("RPT+INH", seq2(.001, .5), seq2(.0001, .1))
mLL_TMC <- calc_mLL("TMC", seq2(.08, .2), seq2(.0001, .15))
```

A function that draws the minus log-likelihood surfaces:

```{r show_mLL}
show_mLL <- function(x, ...) {
  with(x, {
    image(mu_val, sigma_val, mLL_val,
          xlab = expression(paste("within-host mean decay rate ", mu, " (/day)")),
          ylab = expression(paste("individual variability ", sigma, " (/day)")))
    contour(mu_val, sigma_val, mLL_val, add = TRUE,
            levels = c(.002, .01, .1, 1:10), ...)
    box(bty = "o")
  })
}
```

Let's try it:

```{r dependson = "show_mLL", fig.height = 1.75 * 5, fig.width  = 1.5 * 5}
#| label: fig-likelihood-surface
#| fig-cap: "likelihood surface of the model with the data from the 4 treatments of
#| Table 2 of [10.1164/rccm.201103-0397OC](https://doi.org/10.1164/rccm.201103-0397oc)."

opar <- par(mfrow = c(3, 2), cex = 1, plt = c(.17, .93, .18, .89))

walk2(list(mLL_INH, mLL_RIF, mLL_RIFINH, mLL_RPTINH, mLL_TMC),
      c("INH", "RIF", "RIF+INH", "RPT+INH", "TMC"),
      ~ {show_mLL(.x); title2(.y)})

par(opar)
```

### MLE by BFGS

A function that performs estimations using the BFGS quasi-Newton algorithm:

```{r}
estimations <- function(mu, sigma, lower, upper, drug, N0 = 1e4, epsilon = 1e-16) {
  data <- filter(table2, group == drug)
  mle2(function(mu, sigma) mLL(mu, sigma, N0, data, epsilon),
       list(mu = mu, sigma = sigma), "L-BFGS-B",
       lower = lower, upper = upper)
}
```

Let's use it:

```{r warning = FALSE}
est <- list(
  RIF     = estimations(          mu = .08, sigma = .02,
                        lower = c(mu = .06, sigma = .01),
                        upper = c(mu = .09, sigma = .05),
                        drug  = "RIF"),
  RIF_INH = estimations(          mu = .11, sigma = .04,
                        lower = c(mu = .09, sigma = .02),
                        upper = c(mu = .11, sigma = .05),
                        drug  = "RIF+INH"),
  RPT_INH = estimations(          mu = .3 , sigma = .04,
                        lower = c(mu = .28, sigma = .01),
                        upper = c(mu = .32, sigma = .08),
                        drug  = "RPT+INH"),
  TMC     = estimations(          mu = .12, sigma = .04,
                        lower = c(mu = .11, sigma = .03),
                        upper = c(mu = .13, sigma = .06),
                        drug  = "TMC"))
```

Which gives:

```{r}
map_dfr(est, coef, .id = "group")
```

### Model predictions

The following figure shows the estimated inter-individual variability of the
exponential decay rate:

```{r}
#| label: fig-inferred-distribution-decay-rate
#| fig-cap: "inferred distributions of the rate of CFU decay for 3 treatments of Table
#| 2 of [10.1164/rccm.201103-0397OC](https://doi.org/10.1164/rccm.201103-0397oc)."

nb_pts <- 512
x_max <- .45
x_val <- seq(0, x_max, le = nb_pts)
y2 <- rep(0, nb_pts)

add_area <- function(x, col) {
  y1 <- dgamma2(x_val, x["mu"], x["sigma"])
  polygon2(x_val, y1, y2, col = col)
  lines2(x_val, y1, col = col)
}

plot(NA, xlim = c(0, x_max), ylim = c(0, 25),
     xlab = expression(paste("decay rate ", lambda, " (/day)")),
     ylab = "density of probability")

est |> 
  map(coef) |> 
  walk2(col_treatments[-1], add_area)

legend2("topright", legend = treatments[-1], col = col_treatments[-1], lwd = 2)
```

A function that adds model predictions to a plot:

```{r add_pred}
add_pred <- function(x, col, t_max, cfu0 = 1e4, alpha = .2, aggregation = 7) {
  t_val <- seq3(0, t_max)
  t_val2 <- t_val / aggregation
  polygon2(t_val2,
           cfu(cfu0, qgamma2(.025, x["mu"], x["sigma"]), t_val),
           cfu(cfu0, qgamma2(.975, x["mu"], x["sigma"]), t_val), col = col)
  lines2(t_val2, cfu(cfu0, qgamma2(.500, x["mu"], x["sigma"]), t_val), col = col)
}
```

The following function shows the model-predicted expected proportions of
positive individuals in the population for a given treatment:

```{r dependson = "add_pred"}
predicted_cfu2 <- function(est, col, t_max = 6 * 30, agg = 7, log = FALSE,
                           cfu0 = 1e4, ...) {
  plot_frame(t_max = t_max, agg = agg, log = log, ...)
  add_pred(coef(est), col, t_max, cfu0, aggregation = agg)
}
```

This function is a wrapper around `predicted_cfu2()` that calls  `predicted_cfu2()` in
3 different ways:

```{r}
plot1drug_pred <- function(est, col, ttl, ...) {
  predicted_cfu2(est, col, xlab = "time (weeks)", ...)
  predicted_cfu2(est, col, 30, 1, xlab = "time (days)", ...)
  title3(ttl)
  predicted_cfu2(est, col, xlab = "time (weeks)", log = TRUE, ...)
}
```

Let's show the model prediction for all the estimated treatment effects:

```{r fig.width = .7 * 1.7 * 5, fig.height = .7 * 4 * 3.1}
#| label: fig-predicted-cfu-decay
#| fig-cap: "model-predicted decay of CFU from the experimental data of Table 2 of
#| [10.1164/rccm.201103-0397OC](https://doi.org/10.1164/rccm.201103-0397oc)."

opar <- par(mfrow = c(4, 3), cex = 1, plt = c(.24, .99, .2383, .9147))
pwalk(list(est, col_treatments[-1], treatments[-1]),
      plot1drug_pred, mgp = c(1.25, 0, 0))
par(opar)
```

We can compare this model predictions at individual level with Figure 3 of
[10.1164/rccm.201103-0397OC](https://doi.org/10.1164/rccm.201103-0397oc)
that shows CFU after 1 month of treatment, see @fig-figure-3 below.

![Figure 3 from [10.1164/rccm.201103-0397OC](https://doi.org/10.1164/rccm.201103-0397oc).](figure 3.png){#fig-figure-3 width=45%}

A function that adds model-predicted proportions of individual with CFU > 1 in 
a population of 15 individuals:

```{r}
add_mod_pred <- function(x, col, alpha = .2) {
  t_val <- seq3(0, 6 * 30)
  p_val <- map_dbl(t_val, ~ probability(x["mu"], x["sigma"],
                                        N0 = 1e4, t = .x, epsilon = 1e-16))
  t_val <- t_val / 30
  polygon2(t_val,
           qbinom(.025, 15, p_val) / 15,
           qbinom(.975, 15, p_val) / 15, col = col)
  lines2(t_val, qbinom(.5, 15, p_val) / 15, col = col)
}
```

Another version of the `plot_data_prop_est()` function that plot only 1 treatment:

```{r plot_data_prop_est2, dependson = "plot_prop_est"}
plot_data_prop_est2 <- function(ttmt, col, polyg = TRUE, connect = TRUE) {
  plot(NA, xlim = c(0, 6), ylim = 0:1,
     xlab = "duration of treatment (months)",
     ylab = "proportion of positive mice")

  table2a |> 
    filter(group == ttmt) |> 
    plot_prop_est(col, polyg, connect)
}
```

Model-predicted proportion of positive individuals as a function of time in a population
of 15 individuals and in comparison with experimental data:

```{r dependson = "plot_data_prop_est", fig.height = 1.3 * 5, fig.width  = 1.5 * 5}
#| label: fig-data-model-prediction
#| fig-cap: "visual assessment of the fit of the model to the data of Table 2 of
#| [10.1164/rccm.201103-0397OC](https://doi.org/10.1164/rccm.201103-0397oc)."

opar <- par(mfrow = c(2, 2), cex = 1, plt = c(.17, .93, .2, .87))
for (i in 2:5) {
  plot_data_prop_est2(treatments[i], col_treatments[i], FALSE, FALSE)
  add_mod_pred(coef(est[[i - 1]]), col_treatments[i])
  title2(treatments[i])
}
par(opar)
```


## Speculations in humans

Another version of the `predicted_cfu()` function:

```{r predicted_cfu2, dependson = "add_pred"}
predicted_cfu2 <- function(t_max = 6 * 30, agg = 7, log = FALSE,
                           cfu0 = 10^(2:4), est_drug, col, ...) {
  plot_frame(t_max = t_max, agg = agg, log = log, ...)
  walk(cfu0, add_pred,
       x = coef(est_drug), col = col, t_max = t_max, alpha = .2, aggregation = agg)
}
```

Let's see what happens if we change the initial CFU:

```{r dependson = "plot_data_prop_est", fig.height = 1.3 * 5, fig.width  = 1.5 * 5}
#| label: fig-effect-of-initial-cfu2
#| fig-cap: "another way to look at the decay dynamics of CFU as a function of initial
#| CFU value."

opar <- par(mfrow = c(2, 2), cex = 1, plt = c(.17, .93, .2, .87))

for (i in 2:5) {
  predicted_cfu2(est_drug = est[[i - 1]], col = col_treatments[i],
                 log = TRUE, xlab = "time (weeks)")
  title2(treatments[i])
}

par(opar)
```

The following function calculates the time of sterilization:

```{r}
cfu1_time <- function(x, cfu0 = 1e4, p = .5) {
  optimize(
    function(t) abs(log(cfu(cfu0, qgamma2(p, x["mu"], x["sigma"]), t))),
    c(0, 200)
  )$minimum
}
```

The following function calculates the time of sterilization together with
confidence interval:

```{r}
CFU0_effect_cfu1_time <- function(x, cfu0) {
  map_dfc(c(upper = .025, median = .5, lower = .975), cfu1_time, x = x, cfu0 = cfu0)
}
```


```{r}
add_speculation <- function(est, col, alpha = .2,
                            v_val = unlist(map(10^(0:3), ~ .x * 1:10)),
                            h_val = 10 * 0:20) {
  abline2(v = v_val); abline2(h = h_val)
  axis(1); axis(2)
  cfu0_val <- 10^seq3(0, 4)
  out <- map_dfr(cfu0_val, CFU0_effect_cfu1_time, x = coef(est))
  polygon2(cfu0_val, out$upper, out$lower, col = col)
  lines2(cfu0_val, out$median, col = col)
}
```


```{r fig.width = .7 * 1.7 * 5, fig.height = .7 * 4 * 3.1}
#| label: fig-sterilization-and-initial-cfu
#| fig-cap: "time at which sterilization is reached as a function of the initial CFU
#| value."

opar <- par(mfrow = c(4, 3), cex = 1, plt = c(.24, .99, .2383, .9147))

plot_tmp <- function(...) {
  plot(..., axes = FALSE, xlab = "initial CFU", ylab = "time of clearance (days)")
}

for (i in 2:5) {
  plot_tmp(NA, xlim = 10^c(0, 4), ylim = c(0, 200), log = "x")
  add_speculation(est[[i - 1]], col_treatments[i])
  
  plot_tmp(NA, xlim = c(1, 50), ylim = c(0, 100))
  add_speculation(est[[i - 1]], col_treatments[i], v_val = seq(0, 50, 5))
  title3(treatments[i])
  
  plot_tmp(NA, xlim = c(1, 10), ylim = c(0, 50))
  add_speculation(est[[i - 1]], col_treatments[i], h_val = seq(0, 50, 5))
}

par(opar)
```
